A Linear Solution of the Linearity Challenge

### 2025-12-24 paper/Process.lagda.tex
revised processes

All the constructors representing input/output actions, namely those from
\AgdaInductiveConstructor{fail} through \AgdaInductiveConstructor{contract},
basically follow the same pattern: they possibly quantify over some types $A$
and $B$ and then (implicitly) over a typing context $\Context$ through the
function \AgdaFunction{∀[\_]} applied to a predicate of the form
$P~\AgdaFunction{⇒}~\AgdaDatatype{Proc}$. The predicate states how to build a
process that is well typed in $\Context$, provided that $\Context$ satisfies
$P$. In general, $P$ is a (separating) conjunction of sub-predicates
corresponding to the channel on which the process is acting and to the premises
of its typing rule.

For example, the constructor \AgdaInductiveConstructor{fail}, which corresponds
to a process of the form $\Fail\x$, requires the context $\Context$ to satisfy
the predicate
$\AgdaDatatype{Ch}~\AgdaInductiveConstructor{⊤}~\AgdaDatatype{∗}~\AgdaFunction{U}$,
meaning that $\Context$ must contain an entry \AgdaInductiveConstructor{⊤}. That
form of $\Context$ matches the one in the conclusion of \FailRule.

@@

Let us move on to the processes that produce a continuation channel. For
example, the constructor \AgdaInductiveConstructor{case} corresponds to a
process of the form $\Case\x\y{P}{Q}$, where both $P$ and $Q$ may use the
continuation channel $y$. In this case $\Context$ must satisfy the predicate
\[
  \AgdaDatatype{Ch}~(A~\AgdaInductiveConstructor{\&}~B)~\AgdaDatatype{∗}~((A~\AgdaInductiveConstructor{∷\_})~\AgdaFunction{⊢}~\AgdaDatatype{Proc}~\AgdaFunction{∩}~(B~\AgdaInductiveConstructor{∷\_})~\AgdaFunction{⊢}~\AgdaDatatype{Proc})
\]
which looks intimidating at first but makes perfect sense once we recall
\CaseRule. Remember that we are trying to establish whether $\Case\x\y{P}{Q}$ is
well typed in $\Context$.
%
The predicate $\AgdaDatatype{Ch}~(A~\AgdaInductiveConstructor{\&}~B)$ expresses
the requirement that the type of $x$ must be of the form $A \With B$ and should
be found in $\Context$. In other words, it must be the case that
$\Splitting\Context{[A \With B]}\ContextD$ for some $\ContextD$. Now $P$ and $Q$
must be well typed in the context $\ContextD$ augmented with the association $y
: A$ and $y : B$, respectively, whence the use of \AgdaFunction{∩} to verify a
(non-separating) conjunction of the predicates
$(A~\AgdaInductiveConstructor{∷\_})~\AgdaFunction{⊢}~\AgdaDatatype{Proc}$ and
$(B~\AgdaInductiveConstructor{∷\_})~\AgdaFunction{⊢}~\AgdaDatatype{Proc}$. The
two new contexts are obtained by \emph{adding} either $A$ or $B$ to $\ContextD$,
which we perform using \AgdaFunction{⊢}.
%
Crucially, the types $A$ and $B$ are \emph{prepended} to $\ContextD$, which is
consistent with the fact that $y$ has been freshly introduced and $x$ has been
consumed. Notice how easy it is to \emph{prepend} either $A$ or $B$ to
$\ContextD$ instead of \emph{changing} the type of $x$ in $\Context$ from $A
\With B$ to either $A$ or $B$ while preserving the position of the type, as we
would have to do in a ``true'' session type system without explicit continuation
channels.

The interpretation of the remaining constructors is analogous, so we only
comment \AgdaInductiveConstructor{all}, which represents a polymorphic process
$\Forall\x\X\y.P$. This constructor models the continuation $P$ using
higher-order abstract syntax (HOAS~\cite{}): witnessing that a context
$\Context$ satisfies the predicate
\[
  \AgdaFunction{⋂[}~X~\AgdaFunction{∶}~\AgdaDatatype{Type}~\AgdaFunction]~((\AgdaFunction{subst~[}~X~\AgdaFunction{/]}~A~\AgdaInductiveConstructor{∷\_})~\AgdaFunction{⊢}~\AgdaDatatype{Proc}
\]
means providing a function that, for every type $X$, produces a witness for the
predicate
\[
  ((\AgdaFunction{subst~[}~X~\AgdaFunction{/]}~A~\AgdaInductiveConstructor{∷\_})~\AgdaFunction{⊢}~\AgdaDatatype{Proc}
\]
applied to $\Context$. Note that the side condition $X \not\in \fv\Context$ of
\ForallRule is enforced \emph{by definition}:
%
$A$ has type $\AgdaDatatype{PreType}~\AgdaInductiveConstructor{1}$, that is a
pre-type with \emph{at most one} free type variable ($X$), whereas $\Context$ is
a typing context, that is a list of $\AgdaDatatype{Type} =
\AgdaDatatype{PreType}~\AgdaInductiveConstructor{0}$ without free type
variables. Therefore, $X$ cannot occur in $\Context$.

We conclude this module proving that permutations preserve process typing. Since
list permutations basically correspond to channel renaming, we can read this
property as the fact that typing is preserved by (injective) name
substitutions.\Luca{Abbiamo detto da qualche parte dove emergono gli indici di
de Bruijn e di fare attenzione al fatto che lo splitting li altera? Serve questa
digressione?}\Luca{Da qualche parte mettere un remark per far notare che l'uso
della separating conjunction introduce rumore, ma avere i costruttori
currificati ne complica un po' l'interpretazione dei tipi. Forse è meglio fare
questo rilievo nella prossima sezione quando è evidente il pattern matching che
serve per esprimere le regole.}

>>>>

All the constructors except \AgdaInductiveConstructor{link} and
\AgdaInductiveConstructor{cut} basically follow the same pattern: they possibly
quantify over some types $A$ and $B$ and then (implicitly) over a typing context
$\Context$ through the function \AgdaFunction{∀[\_]} applied to a predicate of
the form $P~\AgdaFunction{⇒}~\AgdaDatatype{Proc}$. The predicate states how to
build a process that is well typed in $\Context$, provided that $\Context$
satisfies $P$. In general, $P$ is a (separating) conjunction of sub-predicates
corresponding to the channel on which the process is acting and to the premises
of its typing rule.

For example, the constructor \AgdaInductiveConstructor{fail}, which corresponds
to a process of the form $\Fail\x$, requires the context $\Context$ to satisfy
the predicate
$\AgdaDatatype{Ch}~\AgdaInductiveConstructor{⊤}~\AgdaDatatype{∗}~\AgdaFunction{U}$,
meaning that $\Context$ must contain an entry \AgdaInductiveConstructor{⊤}
(\AgdaFunction{U} is the universal predicate that holds for every context, \cf
\Cref{tab:predicates}). That form of $\Context$ matches the one in the
conclusion of \FailRule.

@@

Let us move on to the processes that produce a continuation channel. For
example, the constructor \AgdaInductiveConstructor{case} corresponds to a
process of the form $\Case\x\y{P}{Q}$, where both $P$ and $Q$ may use the
continuation channel $y$. In this case $\Context$ must satisfy the predicate
\[
  \AgdaDatatype{Ch}~(A~\AgdaInductiveConstructor{\&}~B)~\AgdaDatatype{∗}~((A~\AgdaInductiveConstructor{∷\_})~\AgdaFunction{⊢}~\AgdaDatatype{Proc}~\AgdaFunction{∩}~(B~\AgdaInductiveConstructor{∷\_})~\AgdaFunction{⊢}~\AgdaDatatype{Proc})
\]
which looks intimidating at first but makes perfect sense once we recall the
typing rule \CaseRule and the definitions of \AgdaFunction{∩} and
\AgdaFunction{⊢} in \Cref{tab:predicates}. Remember that we are trying to
establish whether $\Case\x\y{P}{Q}$ is well typed in $\Context$.
%
The predicate $\AgdaDatatype{Ch}~(A~\AgdaInductiveConstructor{\&}~B)$ expresses
the requirement that the type of $x$ must be of the form $A \With B$ and should
be found in $\Context$. In other words, it must be the case that
$\Splitting\Context{[A \With B]}\ContextD$ for some $\ContextD$. Now $P$ and $Q$
must be well typed in the context $\ContextD$ augmented with the association $y
: A$ and $y : B$, respectively, whence the use of \AgdaFunction{∩} to verify a
(non-separating) conjunction of the predicates
$(A~\AgdaInductiveConstructor{∷\_})~\AgdaFunction{⊢}~\AgdaDatatype{Proc}$ and
$(B~\AgdaInductiveConstructor{∷\_})~\AgdaFunction{⊢}~\AgdaDatatype{Proc}$. The
two new contexts are obtained by \emph{adding} either $A$ or $B$ to $\ContextD$,
which we perform using \AgdaFunction{⊢}.
%
Crucially, the types $A$ and $B$ are \emph{prepended} to $\ContextD$, which is
consistent with the fact that the continuation $y$ has been freshly introduced
and the channel $x$ has been consumed. Notice how easy it is to \emph{prepend}
either $A$ or $B$ to $\ContextD$ instead of \emph{changing} the type of $x$ in
$\Context$ from $A \With B$ to either $A$ or $B$ while preserving the position
of the type, as we would have to do in a ``true'' session type system without
explicit continuation channels.

The interpretation of the remaining constructors is analogous, so we only
comment \AgdaInductiveConstructor{all}, which represents a polymorphic process
$\Forall\x\X\y.P$. This constructor models the continuation $P$ using
higher-order abstract syntax: witnessing that a context $\Context$ satisfies the
predicate
\[
  \AgdaFunction{⋂[}~X~\AgdaFunction{∶}~\AgdaDatatype{Type}~\AgdaFunction]~((\AgdaFunction{subst~[}~X~\AgdaFunction{/]}~A~\AgdaInductiveConstructor{∷\_})~\AgdaFunction{⊢}~\AgdaDatatype{Proc}
\]
means providing a function that, for every type $X$, produces a witness for the
predicate
\[
  ((\AgdaFunction{subst~[}~X~\AgdaFunction{/]}~A~\AgdaInductiveConstructor{∷\_})~\AgdaFunction{⊢}~\AgdaDatatype{Proc}
\]
applied to $\Context$. Note that the side condition $X \not\in \fv\Context$ of
\ForallRule is enforced \emph{by definition}:
%
$A$ has type $\AgdaDatatype{PreType}~\AgdaInductiveConstructor{1}$, that is a
pre-type with \emph{at most one} free type variable ($X$), whereas $\Context$ is
a typing context, that is a list of $\AgdaDatatype{Type} =
\AgdaDatatype{PreType}~\AgdaInductiveConstructor{0}$ without free type
variables. Therefore, $X$ cannot occur in $\Context$.

We conclude this module proving that permutations preserve process typing. Since
list permutations basically correspond to channel renaming, we can read this
property as the fact that typing is preserved by (injective) name
substitutions.\Luca{Da qualche parte mettere un remark per far notare che l'uso
della separating conjunction introduce rumore, ma avere i costruttori
currificati ne complica un po' l'interpretazione dei tipi. Forse è meglio fare
questo rilievo nella prossima sezione quando è evidente il pattern matching che
serve per esprimere le regole.}

### 2025-12-24 paper/introduction.tex
revised processes

The rest of the paper is organized as follows.
%
\Cref{sec:calculus} describes the syntax and the operational semantics of \LCP
and states the properties of well-typed processes that we formalize and prove,
namely typing preservation, deadlock freedom and runtime safety.
%
\Cref{sec:agda} presents a roadmap of the Agda formalization and discusses its
key aspects with particular emphasis on the representation of channel names and
of typing contexts.
%
\Cref{sec:examples} illustrates a few examples of \LCP processes as they are
represented in Agda and shows how to use Agda as an interpreter for \LCP.
%
\Cref{sec:related} discusses related work more in detail by providing a
qualitative and quantitative comparison between the known formalizations of
session/linear calculi, including the one for \LCP.
%
\Cref{sec:conclusion} summarizes our contributions and discusses some ongoing
and future work.

>>>>

The rest of the paper is organized as follows.
%
\Cref{sec:calculus} describes the syntax and the operational semantics of \LCP
and states the properties of well-typed processes that we formalize and prove,
namely typing preservation, deadlock freedom and runtime safety.
%
\Cref{sec:agda} presents a roadmap of the Agda formalization and discusses its
key aspects with particular emphasis on the representation of channel names and
of typing contexts. We assume that the reader is somewhat knowledgeable about
Agda, but we also recall a few key definitions from Agda's standard library.
%
\Cref{sec:examples} illustrates some examples of \LCP processes as they are
represented in Agda and shows how to use Agda as an interpreter for \LCP.
%
\Cref{sec:related} discusses related work more in detail by providing a
qualitative and quantitative comparison between the known formalizations of
session/linear calculi, including the one for \LCP.
%
\Cref{sec:conclusion} summarizes our contributions and discusses some ongoing
and future work.
